---
title: "Finding Pythagorean Triples by Brute Force Search"
author: "Jose Torres Postigo"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: false
    extra_dependencies: ["placeins"]
papersize: a4
fontsize: 11pt
geometry:
  - top=1in
  - bottom=1in
  - left=1.5in
  - right=1.5in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Objectives

This lab session outlines the problem to implement the brute-force algorithm to find all primitive Pythagorean triples, given a positive integer $l \in \mathbb{N}$. The objectives of this are not only implement it successfully with improvements, but also to confirm the theoretical number of triples and evaluate the running time of the algorithm. I will accomplish this just by verifying empirically the number of primitive Pythagorean triples, and checking the experimental complexity of the algorithm.

A Pythagorean triple consists of three positive integers $a$, $b$ and $c$, such that $a^2 + b^2 = c^2, \ a<b<c$. This triple is represented like $(a,b,c)$. A primitive Pythagorean triple is one which $a$ and $b$ are coprime, i.e., they have no common divisor larger than 1.

The main problem of this brute-force algorithm is that the complexity is $O(l^3)$ in the first approach because of its inner loops. It can be reduced to $O(l^2)$ with some improvements, so I am going to run the experiment with this last one so this experiment is feasible.



# Experimental Setup

The experiment is conducted easily: I run the algorithm giving the parameter $l$ of the algorithm a very high value: 100\,000 in this case. This will lead me to know better how efficient is this algorithm with these kind of values.

```{=tex}
\begin{table}[!h]
\caption{Computational environment considered.}
\begin{tabular}{lp{0.8\linewidth}}
\hline
CPU       & Intel® Core™ i7-6700HQ × 8, 16.0 GiB \\
OS        & Fedora Linux 38 (Workstation Edition) \\
Java      & openjdk 17.0.8 2023-07-18\\
\hline
\end{tabular}
\label{tab:conf}
\end{table}
```
# Empirical Results

A summary of the experimental results is provided in Table \@ref(tab:data-summary) in the Appendix, along with the statistical fitting of the data to different growth models.

The graph Figure \@ref(fig:numtriples) shows us that the total of pythagorean triples according to the value of $l$, the maximum value, it is correct in every case. This means that the algorithm implemented obtains every primitive Pythagorean triple that can under the value of $l$.

The graph Figure \@ref(fig:time) shows the fit of the data taken doing the experiment to the model. The experiment, as we can see, goes from a maximum value of 10 up to 100\,000; this value is $l$. The figures are given in seconds.

Overall, there is a good trend at the beginning of the graph, from $l=10$ up to $l=50\,000$ approximately. From this value, the fit of the data start being loose regarding the model, and start taking the shape of another one. In this case, looks like is an hyperbolic tangent.

```{r}
# Reading the data
data <- data.frame(read.table("NaivePythagoreanTripleFinder-stats.txt"));
names(data) <- c("maxval", "numtriples", "time")

# Power-law fit

n <- dim(data)[1];
max_x <- 10^(ceiling(log10(max(data$maxval))))
xx <- seq(min(data$maxval), max_x, by = 1)

b0 <- log(data$time[n]/data$time[n-1])/log(data$maxval[n]/data$maxval[n-1])
a0 <- data$time[n]/(data$maxval[n]^b0)
datafit_time <- nls( time ~ a * maxval^b, data=data, start = list(a = a0, b = b0), control=nls.control(maxiter=1000))
fitfun_time <- function(x) predict(datafit_time, newdata=data.frame(maxval=x))

b0 <- log(data$numtriples[n]/data$numtriples[n-1])/log(data$maxval[n]/data$maxval[n-1])
a0 <- data$numtriples[n]/(data$maxval[n]^b0)
datafit_num <- nls(numtriples ~ a * maxval^b, data=data, start = list(a = a0, b = b0))
fitfun_num <- function(x) predict(datafit_num, newdata=data.frame(maxval=x))



pwrdata <- data.frame(
  "maxval" = xx,
  "ptime"  = fitfun_time(xx),
  "pnum"  = fitfun_num(xx)
)
  


```

```{r numtriples, fig.cap="Number of primitive Pythagorean triples up to a maximum value"}
# Graphical representation

library(ggplot2)
library(scales)
col <- c("data" = "blue", "power-law fit" = "red")
rx <-  seq(0, max_x, by = max_x/10)
figure <- ggplot(data=data, aes(x=maxval, y=numtriples)) + 
  geom_line(data = pwrdata, aes(x=maxval, y=pnum, colour = "power-law fit")) +
  geom_point(shape=21, size=3, aes(colour="data"),  fill="white") +
  scale_colour_manual(name = "", values = col, 
                      labels = c("data", 
                                   bquote(.(summary(datafit_num)$coefficients[1]) ~ n^.(summary(datafit_num)$coefficients[2])))
    ) +  
  guides(color = guide_legend(override.aes = list(shape = c(1, NA), 
                                                  linetype = c("blank", "solid")
                                                  ) 
                              )
         ) +
  scale_x_continuous (name = "maximum value", breaks = rx, 
                      labels = scales::number_format(accuracy = 1)) +
  scale_y_continuous(name = "#triples", labels = scales::number_format(accuracy = 1))+
  theme_bw() + 
  theme(legend.justification = c(0, 1), 
        legend.position = c(0, 1), 
        legend.box.margin=margin(c(5,5,5,5))) 

show(figure)
```

```{r time, fig.cap="Time required for finding primitive Pythagorean triples up to a maximum value"}
# Graphical representation

rx <-  seq(0, max_x, by = max_x/10)
figure <- ggplot(data=data, aes(x=maxval, y=time)) + 
  geom_line(data = pwrdata, aes(x=maxval, y=ptime, colour = "power-law fit")) +
  geom_point(shape=21, size=3, aes(colour="data"),  fill="white") +
  scale_colour_manual(name = "", values = col, 
                      labels = c("data", 
                                   bquote(.(summary(datafit_time)$coefficients[1]) ~ n^.(summary(datafit_time)$coefficients[2])))
    ) +
  guides(color = guide_legend(override.aes = list(shape = c(1, NA), 
                                                  linetype = c("blank", "solid")
                                                  ) 
                              )
         ) +
  scale_x_continuous (name = "maximum value", breaks = rx, 
                      labels = scales::number_format(accuracy = 1)) +
  scale_y_continuous(name = "time (s)", labels = scales::number_format(accuracy = 1))+
  theme_bw() + 
  theme(legend.justification = c(0, 1), 
        legend.position = c(0, 1), 
        legend.box.margin=margin(c(5,5,5,5))) 

show(figure)
```

\FloatBarrier

# Discussion

Even though in the graph Figure \@ref(fig:numtriples) I have got a perfect fit, in the graph Figure \@ref(fig:time) the model does not look like is the correct one. Maybe is not this the actual problem, but the hardware i am using to accomplish this experiment.

In my opinion, this model should be reviewed and the data contrasted with another models to find a better one.

This is not the most efficient approach for finding Pythagorean triples, but the brute-force approach used in this project is the easiest to implement. Looking for more approaches to this problem, I found the Euclid's formula.

The Euclid's formula provides a more efficient solutions with a time complexity of $O(n)$. Seems to be the preferred choice for data scientists and software engineers due to its speed and scalability.

\FloatBarrier

```{=tex}
\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}
```
# Appendix

## Data Summary

```{r data-summary}
library(knitr)

table_dataframe <- data.frame(
    "maxval" = data$maxval,
    "num" = data$numtriples,
    "time" = data$time
  )
kable(table_dataframe, "pipe",
      col.names = c("maxval", "#triples", "time"), 
      caption = "Summary of the experimental results")
```

## Model Fitting

### Number of Pythagorean triples

```{r}
show(summary(datafit_num))
```

### Computational time

```{r}
show(summary(datafit_time))
```
